---
title: "A Monte Carlo Analysis of the Generalized Monty Hall Problem"
description: "Project 3: Quantifying the probabilistic advantage of door-switching in an n-door extension of the classic paradox."
author: "Lilian Hu"
date: "March 31, 2025"
format: html
code-copy: true
---

## Theoretical Framework

### Problem Specification

The Monty Hall problem presents a statistical paradox with the following structure:

1.  **Initial Setup**:
    -   A game with $n$ doors (classically $n=3$, generalized here to $n=100$)
    -   One door hides a desirable prize
    -   $n-1$ doors conceal goats (undesired outcomes)
2.  **Game Sequence**:
    1.  Player selects one door (uniformly at random)
    2.  Host opens $k$ doors (where $1 \leq k \leq n-2$), always revealing goats
    3.  Player chooses whether to:
        -   Stay with initial choice, or
        -   Switch to one remaining unopened door

### Probability Analysis

#### Definitions

Let: 
- $P_S$ = Probability of winning when switching 
- $P_{NS}$ = Probability of winning when not switching 
- $n$ = Total number of doors 
- $k$ = Number of doors opened by host ($k = n-2$ in standard version)

#### Mathematical Proof

**Case 1: Not Switching** $$
P_{NS} = \frac{1}{n}
$$ *Rationale*: The player's initial choice has an equal $\frac{1}{n}$ chance of being correct, unaffected by subsequent host actions.

**Case 2: Switching** $$
P_S = \frac{n-1}{n} \times \frac{1}{n-k-1}
$$

For the standard scenario where host opens $k=n-2$ doors: $$
P_S = \frac{n-1}{n} \times \frac{1}{1} = \frac{n-1}{n}
$$

*Derivation*: 1. Initial probability of prize being behind a non-selected door: $\frac{n-1}{n}$ 2. Host's action concentrates this probability mass onto the single remaining unopened door

**Dominance Relationship**: For all $n > 2$: $$
EV_{\text{Switch}} > EV_{\text{Stay}}
$$

### Bayesian Interpretation

The host's actions provide information that updates the probability distribution:

1.  **Prior Belief**: $$
    P(\text{Prize behind initial door}) = \frac{1}{n}
    $$

2.  **Posterior Belief** after observing host's action: $$
    P(\text{Prize behind remaining door}) = \frac{n-1}{n}
    $$

*Key Insight*: The host's door-opening is a *non-random* act that reveals information about the system's state

### Simulation Study

```{r}
set.seed(168)
library(tidyverse)
library(purrr)
library(scales)
```

```{r}

simulate_one_game <- function(n_doors) {
  prize <- sample(n_doors, 1)
  initial_choice <- sample(n_doors, 1)
  
  # Host opens all but one losing door
  remaining_doors <- setdiff(1:n_doors, c(initial_choice, prize))
  monty_opens <- sample(remaining_doors, n_doors - 2, replace = FALSE)
  
  # Determine switched choice
  switched_choice <- setdiff(1:n_doors, c(initial_choice, monty_opens))
  
  tibble(
    stay_win = initial_choice == prize,
    switch_win = switched_choice == prize
  )
}

sim_results <- map_df(
  1:10000, 
  ~ simulate_one_game(n_doors = 100),
  .id = "trial"
)

results <- sim_results |> 
  summarize(
    stay_win_rate = mean(stay_win),
    switch_win_rate = mean(switch_win),
    .groups = 'drop'
  )|> 
  mutate(
    theoretical_stay = 1/100,
    theoretical_switch = 99/100
  )
```

### Results Visualization

```{r}
ggplot(results |> pivot_longer(cols = -contains("theoretical")), 
       aes(x = name, y = value, fill = name)) +
  geom_col(width = 0.6) +
  geom_hline(
    aes(yintercept = theoretical_stay), 
    linetype = "dashed", 
    color = "red"
  ) +
  geom_hline(
    aes(yintercept = theoretical_switch), 
    linetype = "dashed", 
    color = "red"
  ) +
  geom_text(
    aes(label = percent(value, accuracy = 0.1)), 
    vjust = -0.5, 
    size = 5
  ) +
  scale_y_continuous(
    labels = percent_format(), 
    limits = c(0, 1),
    expand = expansion(mult = c(0, 0.1))
  ) +
  scale_x_discrete(
    labels = c("Stay Strategy", "Switch Strategy")
  ) +
  labs(
    title = "Monty Hall Problem with 100 Doors",
    subtitle = "Empirical vs. Theoretical Win Rates (10,000 Simulations)",
    y = "Win Probability",
    x = NULL,
    caption = "Dashed lines show theoretical predictions"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```
